<!-- rmarkdown v1 -->
```{r global,fig.path="figure//"}
```
---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The data was obtained as part ot the Peer Assesment 1 from the Git Repo found here at the below USRl. The file is activity.zip which I unzipped to the directory for the git clone.

https://github.com/rdpeng/RepData_PeerAssessment1

Then I set the working directory and read in the file.

The 5 minute intervals are counted using three digits: h:mm where h is the number of hours since the beginning of the day and mm is the number of minutes since the start of the hour. I'd like to number these a bit more intuitively, just going from 0:287 each day. So I'll add a column for that and also for the actual time in HH:MM format.
```{r loaddata,echo=TRUE}
setwd("~/Coursera/Reproducible/RepData_PeerAssessment1-master")
raw <- read.table(file="activity.csv",sep=",",header=TRUE)
augmentedRaw <- raw[order(raw$date,raw$interval),]
augmentedRaw$steps <- as.numeric(augmentedRaw$steps)
augmentedRaw$ivs <- 0:287 # five minute intervals each day
augmentedRaw$ivsecs <- augmentedRaw$ivs * 5 * 60 # convert to seconds
augmentedRaw$time <- format(.POSIXct(augmentedRaw$ivsecs,tz="GMT"), "%H:%M")

noNAs <- augmentedRaw[!is.na(raw$steps),]
```
## Mean steps per day
What is mean total number of steps taken per day? Note that we will ignore the rows with no data

```{r daily_means, echo=TRUE}
dailySteps <- aggregate(x=noNAs[c("steps")],by=list(Group.date=noNAs$date),FUN=sum)
dailySteps$steps 
hist(dailySteps$steps,main="Histogram of Daily Steps",xlab='Steps per Day')

mn <- as.integer(mean(dailySteps$steps))
md <- as.integer(median(dailySteps$steps))
```
The calculated mean is `r mn` and the median is `r md`.
And as a quick check, we'll print the Summary
```{r daily_steps_summary, echo=TRUE}
summary(dailySteps)
```

## Daily activity pattern
What is the average daily activity pattern? We'll plot the means of the intervals.

```{r average_daily_pattern, echo=TRUE}

intervalMeans <- aggregate(x=noNAs[c("steps")],by=list(Group.interval=noNAs$ivs),FUN=mean)

plot(intervalMeans,type="l",xlab='Average Daily Activity Pattern')

maxInterval <- intervalMeans[intervalMeans$steps==max(intervalMeans$steps),1]

```
The interval with the largest average number of steps is: `r maxInterval`.


##Impute values for missing data
Given that we have rows with missing values, we are going to fill the missing data using the means of the corresponding interval. But first, how many missing values are there?
```{r how_many_missing_values, echo=TRUE}
nRaw <- nrow(raw)
nNotNAs <- nrow(noNAs)
nNAs <- nRaw - nNotNAs
```
There are `r nNAs` rows with no step data. Now replace missing values with the means. This should not perturb the observations.

```{r impute_missing_values, echo=TRUE}
library(plyr)
newDS <- augmentedRaw
newDS <- ddply(newDS, 
              .(interval), 
              function(df) {
                df$steps[is.na(df$steps)] <- mean(df$steps, na.rm=TRUE); return(df)
                }
              )

ndailySteps <- aggregate(x=newDS[c("steps")],by=list(Group.date=newDS$date),FUN=sum)
ndailySteps$steps 
hist(ndailySteps$steps,main="Histogram of Daily Steps with Imputed Data",xlab='Steps per Day')

nmn <- as.integer(mean(dailySteps$steps))
nmd <- as.integer(median(dailySteps$steps))

```
The calculated mean of the new data set with the imputed data is `r nmn` and the median is `r nmd`. The mean and median have not changed, as expected.


Now lets look at how the number of steps per day has changed. First, we'll compute this with the augmentedRaw data and we'll subtract that from the values just calculated for the imputed data.

```{r changes_due_to_imputed_values, echo=TRUE}

rdailySteps <- aggregate(x=augmentedRaw[c("steps")],by=list(Group.date=augmentedRaw$date),FUN=sum)
rdailySteps$steps 

ndailySteps$steps - rdailySteps$steps
```
As you can see above, there is no change to the days that had all of the data as shown by the differences being equal to zero. The only changes are for the days that had no data (all NAs). The values for those missing days are shown above in ndailySteps$steps.

##Weekday vs. weekend patterns
Are there differences in activity patterns between weekdays and weekends? First, lets create a factor to identify weekend days from days during the week. We'' convert the date text to a date object and wee if it is either a Saturday or Sunday and mark those as "weekend". Then we'll mark everything else as "weekday". Finally, we'll check the summary to be sure we have the factor with two levels as needed.
```{r weekend_vs_weekday, echo=TRUE}


newDS$weekOrEnd <- 
  ifelse( ((weekdays(as.Date(newDS$date)) == "Saturday") | 
           (weekdays(as.Date(newDS$date)) == "Sunday")),
         "weekend", 
         "weekday")

newDS$weekOrEnd <- as.factor(newDS$weekOrEnd)
summary(newDS$weekOrEnd)

weekday <- newDS[newDS$weekOrEnd=="weekday",]
weekend <- newDS[newDS$weekOrEnd=="weekend",]

weekendMeans <- aggregate(x=weekend[c("steps")],by=list(Group.interval=weekend$ivs),FUN=mean)
weekdayMeans <- aggregate(x=weekday[c("steps")],by=list(Group.interval=weekday$ivs),FUN=mean)
```
Now we have the data we need, so we'll plot them for easy comparison.

```{r plot_diferences, wcho=TRUE}
par(mfrow=c(2,1))
plot(weekdayMeans,type="l",ylab="Average Steps",main="Weekday Average Steps by Interval",xlab='Interval')
plot(weekendMeans,type="l",ylab="Average Steps",main="Weekend Average Steps by Interval",xlab='Interval')


```
